---
title: "Senior thesis"
author: "Ellyse"
date: "3/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
library(tidyverse)
library(janitor)
library(dplyr)
```

```{r}
#read in data
data <- read.csv("mohawk and naples data.csv")
str(data)
#tidy data 
data_tidy <- data %>%
  pivot_longer(`Neighbor.1`:`Neighbor.Extra`) %>% #make each neighbor observation a row
  clean_names() %>% #simplify column names
  mutate(name = str_to_lower(name), already_counted = str_to_lower(already_counted)) #make values in "name" and "already counted?" columns lowercase
#View(data_tidy)

data_tidier <- data_tidy %>% 
  mutate(already_counted=replace(already_counted, already_counted=="no", "NA")) %>% #replace "no" values with NA
  separate(already_counted, c("already_counted_1", "already_counted_2"), "and ") %>% #separate "already counted" values with 2 duplicate observations
  separate(already_counted_1, c("already_counted_1", "already_counted_3","already_counted_4"), ", ") %>% #separate "already counted" values with 3-4 duplicate observations
  mutate(already_counted_2=ifelse(is.na(already_counted_2)==TRUE, "NA", paste0("neighbor ", already_counted_2)),
         already_counted_3=ifelse(is.na(already_counted_3)==TRUE, "NA", paste0("neighbor ", already_counted_3)),
         already_counted_4=ifelse(already_counted_4>0, paste0("neighbor ", already_counted_4), "NA")) %>% #add "neighbor" before each cell containing a viable value
  mutate(already_counted_4=replace_na(already_counted_4, "NA")) %>% #replace implicit NA values with explicit NA values (for filtering)
  filter((already_counted_1!=name)&(already_counted_2!=name)&(already_counted_3!=name)&(already_counted_4!=name)) %>% #remove duplicate observations
  filter(value!="N/A") %>% #remove neighbor observations of "NA"
  mutate(focal = tolower(focal))
#View(data_tidier)
```

Replace shorthand with scientific names
```{r}
sci_name <- read_csv("name key .csv") %>%
  clean_names() %>% 
  mutate(scientific_name = tolower(scientific_name))
#View(sci_name)

data_tidier <- data_tidier %>% 
 rename(shorthand = value) %>% #rename value column to shorthand, for consistency between data frames
  filter(shorthand != "same organism") %>% #remove observations where the neighbor organism is the same as the focal organsim
  mutate(shorthand = ifelse(shorthand=="point occurance corynactis neighbor", "coca", shorthand)) #simplify observed point occurance of corynactis to corynactis

data_name <- merge(data_tidier, sci_name, by='shorthand', all.x=TRUE)
#View(data_name)
```

El nino years 2015- 2016 
```{r}
#creating a column in data_name for el nino event called enso using dataset column 
data_name$enso <- data_name$dataset
#replacing all dates containing 2015 or 2016 with yes and all other years (2013 2104,2107,2018) with no 
data_name$enso[grep("2015|2016", data_name$enso)] <- "yes"
data_name$enso[grep("2013|2014|2017|2018", data_name$enso)] <- "no"
#View(data_name)
```

Seasons
```{r}
#create column in data_name called upwell using dataset column 
data_name$upwell <- data_name$dataset
#replace summer months april-aug (upwelling) with yes and winter months sept-march with no
data_name$upwell[grep("May|June|July|Aug", data_name$upwell)] <- "yes"
data_name$upwell[grep("Jan|Feb|Oct|Nov|Dec", data_name$upwell)] <- "no"
View(data_name)
```

Total neighbor events between coca and urchin in naples and mohawk 
```{r}
#quantify number of times focal= coca and scientific_name= mefr or stpu when location= mohawk
neighbor_mohawk_stpu <- filter(data_name, (focal=="corynactis" & location=="Mohawk" & scientific_name=="echinodermata- strongylocentrotus purpuratus"))
neighbor_mohawk_mefr <- filter(data_name, (focal=="corynactis" & location=="Mohawk" & scientific_name=="echinodermata- mesocentrotus franciscanus"))
#counting the rows in each data frame 
Ms <- nrow(neighbor_mohawk_stpu)
Mm <- nrow(neighbor_mohawk_mefr)
#total number of interactions bewteen coca and urchins at mohawk 
total_int_M <- Ms+Mm #26 

#standardizing interaction by number of photos
photo_mohawk <- sum(data_name$location == "Mohawk")#3623 assign value if more data added 
total_int_M/photo_mohawk #0.007176373 interaction event per Mohawk photo 

```


```{r}
#quantify number of times focal= coca and scientific_name= mefr or stpu when location= naples
neighbor_naples_stpu <- filter(data_name, (focal=="corynactis" & location=="Naples" & scientific_name=="echinodermata- strongylocentrotus purpuratus"))
neighbor_naples_mefr <- filter(data_name, (focal=="corynactis" & location=="Naples" & scientific_name=="echinodermata- mesocentrotus franciscanus"))
#counting the rows in each data frame 
Ns <- nrow(neighbor_naples_stpu)
Nm <- nrow(neighbor_naples_mefr)
#total number of interactions bewteen coca and urchins at naples 
total_int_N <- Ns+Nm #186

#standardizing interaction by number of photos
photo_naples <- sum(data_name$location == "Naples") #10012
total_int_N/photo_naples #0.01857771 interaction events per Naples photo

#what test to use to compare interaction events at two locations? 
```

Counting interaction events during enso vs not 
```{r}
#quantify number of times focal= coca and scientific_name= mefr or stpu when enso= yes 
neighbor_enso_stpu <- filter(data_name, (focal=="corynactis" & scientific_name=="echinodermata- strongylocentrotus purpuratus" & enso=="yes"))
neighbor_enso_mefr <- filter(data_name, (focal=="corynactis" & scientific_name=="echinodermata- mesocentrotus franciscanus" & enso=="yes"))
#counting the rows in each data frame 
Es <- nrow(neighbor_enso_stpu)
Em <- nrow(neighbor_enso_mefr)
#total number of interactions bewteen coca and urchins during enso  
total_int <- Es+Em #81

#standardizing interaction by number of photos
enso_yes <- sum(data_name$enso == "yes") #7757
total_int/enso_yes #0.01044218 interaction events per photo during enso year 

```

```{r}
#quantify number of times focal= coca and scientific_name= mefr or stpu when enso= no
neighbor_ne_stpu <- filter(data_name, (focal=="corynactis" & scientific_name=="echinodermata- strongylocentrotus purpuratus" & enso=="no"))
neighbor_ne_mefr <- filter(data_name, (focal=="corynactis" & scientific_name=="echinodermata- mesocentrotus franciscanus" & enso=="no"))
#counting the rows in each data frame 
Nes <- nrow(neighbor_ne_stpu)
Nem <- nrow(neighbor_ne_mefr)
#total number of interactions bewteen coca and urchins during no enso  
total_int <- Nes+Nem #131

#standardizing interaction by number of photos
enso_no <- sum(data_name$enso == "no") #5878
total_int/enso_no #0.02228649 interaction events per photo during non enso year 

#standardize these counts by photos looked at in 2015/2016 vs all others? would skew the data too much because two years vs a lot of other years?  
```

Counting interactions during season 
```{r}
#quantify number of times focal= coca and scientific_name= mefr or stpu when upwell= yes 
neighbor_upwell_stpu <- filter(data_name, (focal=="corynactis" & scientific_name=="echinodermata- strongylocentrotus purpuratus" & upwell=="yes"))
neighbor_upwell_mefr <- filter(data_name, (focal=="corynactis" & scientific_name=="echinodermata- mesocentrotus franciscanus" & upwell=="yes"))
#counting the rows in each data frame 
Us <- nrow(neighbor_upwell_stpu)
Um <- nrow(neighbor_upwell_mefr)
#total number of interactions bewteen coca and urchins during upwell 
total_int <- Us+Um #121

#standardizing interaction by number of photos
upwell_yes <- sum(data_name$upwell == "yes") #5522
total_int/upwell_yes #0.02191235 interaction events per photo during upwelling season  
```

```{r}
#quantify number of times focal= coca and scientific_name= mefr or stpu when upwell= no
neighbor_nu_stpu <- filter(data_name, (focal=="corynactis" & scientific_name=="echinodermata- strongylocentrotus purpuratus" & upwell=="no"))
neighbor_nu_mefr <- filter(data_name, (focal=="corynactis" & scientific_name=="echinodermata- mesocentrotus franciscanus" & upwell=="no"))
#counting the rows in each data frame 
Nus <- nrow(neighbor_nu_stpu)
Num <- nrow(neighbor_nu_mefr)
#total number of interactions bewteen coca and urchins during upwell 
total_int <- Nus+Num #91

#standardizing interaction by number of photos
upwell_no <- sum(data_name$upwell == "no") #8113
total_int/upwell_no #0.01121657 interaction events per photo during non upwelling season
```

............................................................................

Chi square test 
need to determine how many interactions occur btw coca and not an urchin and divide by total pictures taken 
```{r}
#counting interactions between coca and not an urchin between naples and mohawk 
N_coca_other  <- filter(data_name, (focal=="corynactis" & !scientific_name=="echinodermata- strongylocentrotus purpuratus" & !scientific_name=="echinodermata- mesocentrotus franciscanus" & location=="Naples"))
N <- nrow(N_coca_other) #8427

M_coca_other  <- filter(data_name, (focal=="corynactis" & !scientific_name=="echinodermata- strongylocentrotus purpuratus" & !scientific_name=="echinodermata- mesocentrotus franciscanus" & location=="Mohawk"))
M <- nrow(M_coca_other) #540

#standardize by total number of photos taken per site 
N/photo_naples #0.84169
M/photo_mohawk #0.1490478
```
```{r}
#chi square test naples and mohawk, Naples/Mohawk Interaction/Not interaction. total counts, not standardized by total number of photos taken because would no longer be count data. Still valid? 

chi1 <- matrix(c(26,186,540,8427),nrow=2,ncol=2)
chisq.test(chi1,correct=FALSE) #p-value= 0.0001881
```

```{r}
#counting interactions btw coca and non urchin enso events
En_coca_other  <- filter(data_name, (focal=="corynactis" & !scientific_name=="echinodermata- strongylocentrotus purpuratus" & !scientific_name=="echinodermata- mesocentrotus franciscanus" & enso=="yes"))
En <- nrow(En_coca_other) #5794

Nen_coca_other  <- filter(data_name, (focal=="corynactis" & !scientific_name=="echinodermata- strongylocentrotus purpuratus" & !scientific_name=="echinodermata- mesocentrotus franciscanus" & enso=="no"))
Nen <- nrow(Nen_coca_other) #3173

#chi square
chi2 <- matrix(c(81,131,5794,3173),nrow=2,ncol=2)
chisq.test(chi2,correct=FALSE) #p-value = 2.424e-15
```

```{r}
#counting interactions btw coca and non urchin season
U_coca_other  <- filter(data_name, (focal=="corynactis" & !scientific_name=="echinodermata- strongylocentrotus purpuratus" & !scientific_name=="echinodermata- mesocentrotus franciscanus" & upwell=="yes"))
U <- nrow(U_coca_other) #3398

Nu_coca_other  <- filter(data_name, (focal=="corynactis" & !scientific_name=="echinodermata- strongylocentrotus purpuratus" & !scientific_name=="echinodermata- mesocentrotus franciscanus" & upwell=="no"))
Nu <- nrow(Nu_coca_other) #5569

#chi square
chi3 <- matrix(c(121,91,3398,5569),nrow=2,ncol=2)
chisq.test(chi3,correct=FALSE) #p-value = 1.368e-08
```


interaction between season and enso event? 
chi square code:
Naples/Mohawk Interaction/Not interaction
Nap <- c(number, number)
Ooh <- c(intx, not)
Chi1 <- chisq.test(nap, ooh) assign a variable 
make table to check work and table input order 
chi1$p.value

check that the numbers are correctly used in table
round(chi1$expected,2)
residuals 
round(chi1$residuals, 3)
plot residuals 
corrplot(chi1$residuals, is.cor = FALSE)

............................................................................

GLM
```{r}
#creating new columns for binomial glm 
#new data frame to make changes to 
glm_data <- data.frame(data_name)
glm_data$enso[grep("yes", glm_data$enso)] <- "1"
glm_data$enso[grep("no", glm_data$enso)] <- "0"
glm_data$upwell[grep("yes", glm_data$upwell)] <- "1"
glm_data$upwell[grep("no", glm_data$upwell)] <- "0"
#View(glm_data)

#ID focal=corynactis and neighbor=urchin then divide by total neighbors for coca
#ifelse statement practice 
glm_data$neighbor <- ifelse(glm_data$shorthand=="mefr","heck yes!","NOWAY") 
#nesting ifelse
glm_data$neighbor2 <- ifelse(glm_data$shorthand=="mefr",
                             ifelse(glm_data$orientation=="vertical", "yay!", "notascool"),"heck no")

glm_data$neighbor3 <- ifelse(glm_data$shorthand=="mefr",
                             ifelse(glm_data$orientation=="vertical", "yay!", "notascool"),glm_data$neighbor)    #replacing with another column                         
sort(unique(glm_data$neighbor3))

View(glm_data)
#if/then function argument. if focal=coca and shorthand= urchin then=1 (otherwise 0) add up the number of ones to get total number interactions 
# finding denominator= nrow= total number of rows 
# glmm(percentage~site+year+sitexyear….) Where response is #urchin interactions/#any neighbor
#  glm(admit ~ gre + rank, data=bindat, family="binomial")
# mixed model https://stats.stackexchange.com/questions/58900/fixed-vs-random-effects
#https://stats.stackexchange.com/questions/4700/what-is-the-difference-between-fixed-effect-random-effect-and-mixed-effect-mode


```





