---
title: "Senior thesis"
author: "Ellyse"
date: "3/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(dplyr)
```

```{r}
#read in data
data <- read.csv("mohawk and naples data.csv")
str(data)
#tidy data 
data_tidy <- data %>%
  pivot_longer(`Neighbor.1`:`Neighbor.Extra`) %>% #make each neighbor observation a row
  clean_names() %>% #simplify column names
  mutate(name = str_to_lower(name), already_counted = str_to_lower(already_counted)) #make values in "name" and "already counted?" columns lowercase
#View(data_tidy)

data_tidier <- data_tidy %>% 
  mutate(already_counted=replace(already_counted, already_counted=="no", "NA")) %>% #replace "no" values with NA
  separate(already_counted, c("already_counted_1", "already_counted_2"), "and ") %>% #separate "already counted" values with 2 duplicate observations
  separate(already_counted_1, c("already_counted_1", "already_counted_3","already_counted_4"), ", ") %>% #separate "already counted" values with 3-4 duplicate observations
  mutate(already_counted_2=ifelse(is.na(already_counted_2)==TRUE, "NA", paste0("neighbor ", already_counted_2)),
         already_counted_3=ifelse(is.na(already_counted_3)==TRUE, "NA", paste0("neighbor ", already_counted_3)),
         already_counted_4=ifelse(already_counted_4>0, paste0("neighbor ", already_counted_4), "NA")) %>% #add "neighbor" before each cell containing a viable value
  mutate(already_counted_4=replace_na(already_counted_4, "NA")) %>% #replace implicit NA values with explicit NA values (for filtering)
  filter((already_counted_1!=name)&(already_counted_2!=name)&(already_counted_3!=name)&(already_counted_4!=name)) %>% #remove duplicate observations
  filter(value!="N/A") %>% #remove neighbor observations of "NA"
  mutate(focal = tolower(focal))
#View(data_tidier)
unique(data_tidier$already_counted_1)

#fixing missspellings 
data_tidier[data_tidier == "neighor 1"] <- "neighbor 1"
data_tidier[data_tidier == "neigbor 1"] <- "neighbor 1"
data_tidier[data_tidier == "neighbor 1 "] <- "neighbor 1"
data_tidier[data_tidier == "neghbor 1"] <- "neighbor 1"
data_tidier[data_tidier == "niehgbor 1 "] <- "neighbor 1"
data_tidier[data_tidier == "neighbor 2 "] <- "neighbor 2"
data_tidier[data_tidier == "neighbor 3 "] <- "neighbor 3"
data_tidier[data_tidier == "nrighbor 1"] <- "neighbor 1"
unique(data_tidier$already_counted_1)
```

Replace shorthand with scientific names
```{r}
sci_name <- read_csv("name key .csv") %>%
  clean_names() %>% 
  mutate(scientific_name = tolower(scientific_name))
#View(sci_name)

data_tidier <- data_tidier %>% 
 rename(shorthand = value) %>% #rename value column to shorthand, for consistency between data frames
  filter(shorthand != "same organism") %>% #remove observations where the neighbor organism is the same as the focal organsim
  mutate(shorthand = ifelse(shorthand=="point occurance corynactis neighbor", "coca", shorthand)) #simplify observed point occurance of corynactis to corynactis

data_name <- merge(data_tidier, sci_name, by='shorthand', all.x=TRUE)
#View(data_name)

#sort(unique(data_tidier$shorthand))
#fix inconsistencies in spelling
data_tidier[data_tidier == "same organisms"] <- "same organism"
data_tidier[data_tidier == "barob"] <- "baro"
data_tidier[data_tidier == "csauu"] <- "csau"
data_tidier[data_tidier == "hso"] <- "hsou"
data_tidier[data_tidier == "ndshs"] <- "ndsh"
data_tidier[data_tidier == "obsp"] <- "ob"
data_tidier[data_tidier == "plsp"] <- "pl"
data_tidier[data_tidier == "same"] <- "same organism"
data_tidier[data_tidier == "uteq"] <- "ureq"
data_tidier[data_tidier == "yesp"] <- "yes"

#sort(unique(data_tidier$shorthand))
```

El nino years 2015- 2016 
```{r}
#creating a column in data_name for el nino event called enso using dataset column 
data_name$enso <- data_name$dataset

#replacing dates March 2015-June 2016 with 1 and all other times with 0
#adding year column 
library(stringr)
data_name$year <- str_extract(data_name$dataset, "2013|2014|2015|2016|2017|2018")
sort(unique(data_name$year))
#adding month column 
data_name$season <- str_extract(data_name$dataset, "Jan|Feb|May|June|July|Aug|Oct|Nov|Dec")
sort(unique(data_name$season))

data_name$enso1 <- ifelse(data_name$year=="2015", 
                      ifelse(data_name$season=="May"| data_name$season=="June"|data_name$season=="July"|data_name$season=="Aug"|data_name$season=="Oct"|data_name$season=="Nov"|data_name$season=="Dec",1,0),0)

data_name$enso2 <- ifelse(data_name$year=="2016", 
                      ifelse(data_name$season=="Jan"| data_name$season=="Feb"|data_name$season=="May"|data_name$season=="June",1,0),0)
#combine if else statements into one column called enso
data_name$enso <- ifelse(data_name$enso1==1|data_name$enso2==1, 1, 0)
#View(data_name)
```

Seasons
```{r}
#create column in data_name called upwell using dataset column 
data_name$upwell <- data_name$dataset
#replace summer months april-aug (upwelling) with yes and winter months sept-march with no
data_name$upwell[grep("May|June|July|Aug", data_name$upwell)] <- "yes"
data_name$upwell[grep("Jan|Feb|Oct|Nov|Dec", data_name$upwell)] <- "no"
#View(data_name)
```

.....................................................................
Start of chi square data 

Total neighbor events between coca and urchin in naples and mohawk 
```{r}
#quantify number of times focal= coca and scientific_name= mefr or stpu when location= mohawk
neighbor_mohawk_stpu <- filter(data_name, (focal=="corynactis" & location=="Mohawk" & scientific_name=="echinodermata- strongylocentrotus purpuratus"))
neighbor_mohawk_mefr <- filter(data_name, (focal=="corynactis" & location=="Mohawk" & scientific_name=="echinodermata- mesocentrotus franciscanus"))
#counting the rows in each data frame 
Ms <- nrow(neighbor_mohawk_stpu)
Mm <- nrow(neighbor_mohawk_mefr)
#total number of interactions bewteen coca and urchins at mohawk 
total_int_M <- Ms+Mm #26 

#standardizing interaction by number of photos
photo_mohawk <- sum(data_name$location == "Mohawk")#3623 assign value if more data added 
total_int_M/photo_mohawk #0.007176373 interaction event per Mohawk photo 

```


```{r}
#quantify number of times focal= coca and scientific_name= mefr or stpu when location= naples
neighbor_naples_stpu <- filter(data_name, (focal=="corynactis" & location=="Naples" & scientific_name=="echinodermata- strongylocentrotus purpuratus"))
neighbor_naples_mefr <- filter(data_name, (focal=="corynactis" & location=="Naples" & scientific_name=="echinodermata- mesocentrotus franciscanus"))
#counting the rows in each data frame 
Ns <- nrow(neighbor_naples_stpu)
Nm <- nrow(neighbor_naples_mefr)
#total number of interactions bewteen coca and urchins at naples 
total_int_N <- Ns+Nm #186

#standardizing interaction by number of photos
photo_naples <- sum(data_name$location == "Naples") #10012
total_int_N/photo_naples #0.01857771 interaction events per Naples photo

#what test to use to compare interaction events at two locations? 
```

Counting interaction events during enso vs not 
```{r}
#quantify number of times focal= coca and scientific_name= mefr or stpu when enso= yes 
neighbor_enso_stpu <- filter(data_name, (focal=="corynactis" & scientific_name=="echinodermata- strongylocentrotus purpuratus" & enso=="yes"))
neighbor_enso_mefr <- filter(data_name, (focal=="corynactis" & scientific_name=="echinodermata- mesocentrotus franciscanus" & enso=="yes"))
#counting the rows in each data frame 
Es <- nrow(neighbor_enso_stpu)
Em <- nrow(neighbor_enso_mefr)
#total number of interactions bewteen coca and urchins during enso  
total_int <- Es+Em #81

#standardizing interaction by number of photos
enso_yes <- sum(data_name$enso == "yes") #7757
total_int/enso_yes #0.01044218 interaction events per photo during enso year 

```

```{r}
#quantify number of times focal= coca and scientific_name= mefr or stpu when enso= no
neighbor_ne_stpu <- filter(data_name, (focal=="corynactis" & scientific_name=="echinodermata- strongylocentrotus purpuratus" & enso=="no"))
neighbor_ne_mefr <- filter(data_name, (focal=="corynactis" & scientific_name=="echinodermata- mesocentrotus franciscanus" & enso=="no"))
#counting the rows in each data frame 
Nes <- nrow(neighbor_ne_stpu)
Nem <- nrow(neighbor_ne_mefr)
#total number of interactions bewteen coca and urchins during no enso  
total_int <- Nes+Nem #131

#standardizing interaction by number of photos
enso_no <- sum(data_name$enso == "no") #5878
total_int/enso_no #0.02228649 interaction events per photo during non enso year 

#standardize these counts by photos looked at in 2015/2016 vs all others? would skew the data too much because two years vs a lot of other years?  
```

Counting interactions during season 
```{r}
#quantify number of times focal= coca and scientific_name= mefr or stpu when upwell= yes 
neighbor_upwell_stpu <- filter(data_name, (focal=="corynactis" & scientific_name=="echinodermata- strongylocentrotus purpuratus" & upwell=="yes"))
neighbor_upwell_mefr <- filter(data_name, (focal=="corynactis" & scientific_name=="echinodermata- mesocentrotus franciscanus" & upwell=="yes"))
#counting the rows in each data frame 
Us <- nrow(neighbor_upwell_stpu)
Um <- nrow(neighbor_upwell_mefr)
#total number of interactions bewteen coca and urchins during upwell 
total_int <- Us+Um #121

#standardizing interaction by number of photos
upwell_yes <- sum(data_name$upwell == "yes") #5522
total_int/upwell_yes #0.02191235 interaction events per photo during upwelling season  
```

```{r}
#quantify number of times focal= coca and scientific_name= mefr or stpu when upwell= no
neighbor_nu_stpu <- filter(data_name, (focal=="corynactis" & scientific_name=="echinodermata- strongylocentrotus purpuratus" & upwell=="no"))
neighbor_nu_mefr <- filter(data_name, (focal=="corynactis" & scientific_name=="echinodermata- mesocentrotus franciscanus" & upwell=="no"))
#counting the rows in each data frame 
Nus <- nrow(neighbor_nu_stpu)
Num <- nrow(neighbor_nu_mefr)
#total number of interactions bewteen coca and urchins during upwell 
total_int <- Nus+Num #91

#standardizing interaction by number of photos
upwell_no <- sum(data_name$upwell == "no") #8113
total_int/upwell_no #0.01121657 interaction events per photo during non upwelling season
```

......................................................................

Chi square test 
need to determine how many interactions occur btw coca and not an urchin and divide by total pictures taken 
```{r}
#counting interactions between coca and not an urchin between naples and mohawk 
N_coca_other  <- filter(data_name, (focal=="corynactis" & !scientific_name=="echinodermata- strongylocentrotus purpuratus" & !scientific_name=="echinodermata- mesocentrotus franciscanus" & location=="Naples"))
N <- nrow(N_coca_other) #8427

M_coca_other  <- filter(data_name, (focal=="corynactis" & !scientific_name=="echinodermata- strongylocentrotus purpuratus" & !scientific_name=="echinodermata- mesocentrotus franciscanus" & location=="Mohawk"))
M <- nrow(M_coca_other) #540

#standardize by total number of photos taken per site 
N/photo_naples #0.84169
M/photo_mohawk #0.1490478
```
```{r}
#chi square test naples and mohawk, Naples/Mohawk Interaction/Not interaction. total counts, not standardized by total number of photos taken because would no longer be count data. Still valid? 

chi1 <- matrix(c(26,186,540,8427),nrow=2,ncol=2)
chisq.test(chi1,correct=FALSE) #p-value= 0.0001881
```

```{r}
#counting interactions btw coca and non urchin enso events
En_coca_other  <- filter(data_name, (focal=="corynactis" & !scientific_name=="echinodermata- strongylocentrotus purpuratus" & !scientific_name=="echinodermata- mesocentrotus franciscanus" & enso=="yes"))
En <- nrow(En_coca_other) #5794

Nen_coca_other  <- filter(data_name, (focal=="corynactis" & !scientific_name=="echinodermata- strongylocentrotus purpuratus" & !scientific_name=="echinodermata- mesocentrotus franciscanus" & enso=="no"))
Nen <- nrow(Nen_coca_other) #3173

#chi square
chi2 <- matrix(c(81,131,5794,3173),nrow=2,ncol=2)
chisq.test(chi2,correct=FALSE) #p-value = 2.424e-15
```

```{r}
#counting interactions btw coca and non urchin season
U_coca_other  <- filter(data_name, (focal=="corynactis" & !scientific_name=="echinodermata- strongylocentrotus purpuratus" & !scientific_name=="echinodermata- mesocentrotus franciscanus" & upwell=="yes"))
U <- nrow(U_coca_other) #3398

Nu_coca_other  <- filter(data_name, (focal=="corynactis" & !scientific_name=="echinodermata- strongylocentrotus purpuratus" & !scientific_name=="echinodermata- mesocentrotus franciscanus" & upwell=="no"))
Nu <- nrow(Nu_coca_other) #5569

#chi square
chi3 <- matrix(c(121,91,3398,5569),nrow=2,ncol=2)
chisq.test(chi3,correct=FALSE) #p-value = 1.368e-08

```


interaction between season and enso event? 
chi square code:
Naples/Mohawk Interaction/Not interaction
Nap <- c(number, number)
Ooh <- c(intx, not)
Chi1 <- chisq.test(nap, ooh) assign a variable 
make table to check work and table input order 
chi1$p.value

check that the numbers are correctly used in table
round(chi1$expected,2)
residuals 
round(chi1$residuals, 3)
plot residuals 
corrplot(chi1$residuals, is.cor = FALSE)

......................................................................

GLM
```{r}
#new data frame to use for glm data
glm_data <- data.frame(data_name)
glm_data$upwell[grep("yes", glm_data$upwell)] <- 1
glm_data$upwell[grep("no", glm_data$upwell)] <- 0
#View(glm_data)

#ifelse statement to create new column called neighbor for whenever coca and urchin are next to each other
glm_data$neighbor1 <- ifelse(glm_data$focal=="corynactis", 
                      ifelse(glm_data$shorthand=="mefr" | glm_data$shorthand=="stpu",1,0),0)

glm_data$neighbor2 <- ifelse(glm_data$focal=="mesocentrotus" | glm_data$focal=="strongylocentotus", ifelse(glm_data$shorthand=="coca",1,0),0)

glm_data$neighbor <- ifelse(glm_data$neighbor1==1 | glm_data$neighbor2==1,1,0)

length(which(glm_data$neighbor==1)) #390  
#View(glm_data)


#new column to standardize neighbor by total number of neighbors of coca when enso event 
stand_enso  <- filter(glm_data, (focal=="corynactis" & enso== 1))
E <- nrow(stand_enso) #5943 total neighbors of coca during enso 

#if enso= 1 and neighbor= 1 then divide neighbor interaction by E (the total number of interactions during enso)
glm_data$stand_enso <- ifelse(glm_data$enso==1 & glm_data$neighbor==1, 1/5943, 0)
View(glm_data)



#if/then function argument. if focal=coca and shorthand= urchin then=1 (otherwise 0) add up the number of ones to get total number interactions 
# finding denominator= nrow= total number of rows 
# glmm(percentage~site+year+sitexyear….) Where response is #urchin interactions/#any neighbor
#  glm(admit ~ gre + rank, data=bindat, family="binomial")
# mixed model https://stats.stackexchange.com/questions/58900/fixed-vs-random-effects
#https://stats.stackexchange.com/questions/4700/what-is-the-difference-between-fixed-effect-random-effect-and-mixed-effect-mode

```

```{r}
#GLM test runs 
library(car)
library(MASS)
library(ggplot2)
library(psych)
#GLM mixed effects model 
library(lme4)

#glm practice models
glm1 <- glm(stand_enso ~ year, data=glm_data)
summary(glm1) #2015 and 2016 significant 

```

```{r}
#new column to standardize neighbor by total number of neighbors of coca when seasonal upwelling occurring 
stand_upwell  <- filter(glm_data, (focal=="corynactis" & upwell=="1"))
U <- nrow(stand_upwell) #3600 total neighbors of coca during upwell
sum(U)

#if upwell= 1 and neighbor= 1 then divide neighbor interaction by U (the total number of interactions during upwell)
glm_data$stand_upwell <- ifelse(glm_data$upwell=="1" & glm_data$neighbor=="1", 1/3600, "0")
#View(glm_data)

```

```{r}
#binomial glm for non standardized neighbor events by multiple factors (season, year, location)
glm_data$neighbor <- as.factor(glm_data$neighbor) #not numeric because this is a category, yes/no
glm2 <- glm(neighbor ~ year + season, data=glm_data, family="binomial")
summary(glm2)
glm3 <- glm(neighbor ~ enso + upwell + location, data=glm_data, family="binomial")
summary(glm3)  #enso and location significant predictors 

#previously glm9
glm_data$filename<- as.factor(glm_data$filename)
glm4 <- glmer(neighbor ~ enso + upwell + location + (1|filename) , data=glm_data, family= "binomial")
summary(glm4) #location naples significant predictors of neighbor events 

#previously glm10
glm5 <- glmer(neighbor ~ location + (1|filename), data=glm_data, family="binomial")
summary(glm5) #includes only independent variable that is significant predictors of the model

#test if data fit model assumptions 
#install.packages("DHARMa")
library(DHARMa)
simulationOutput <- simulateResiduals(fittedModel = glm5, plot = T) #gives message to use testOutliers and bootstrap, significant deviation

testOutliers(simulationOutput = glm5, margin = c("both"), type = c("bootstrap"), plot = T) #p value < 0.05 so no overdispersion/outliers
?testOutliers

#install.packages("performance")
library(performance)
#install.packages("see")
library(see)
binned_residuals(glm4) #31% model not a great fit 
testDispersion(glm4)# >0.05, so there is no overdispersion  
binned_residuals(glm5) #32% of the residuals are inside the error bounds. model is a not a good fit  
testDispersion(glm5) # >0.05, so there is no overdispersion  
```

Amelia's addition - checking out binned residuals
```{r}
glm_all <- glmer(neighbor ~ year + upwell + season + location + enso, data=glm_data, family="binomial")
stepAIC(glm_all, direction="both") #for now, this is saying season and location is most predictive (but also, no random effects accounted for becaues stepAIC() can't handle those so take this with a grain of salt)

# with model selection, you also want to think of this from a biological standpoint - should we *actually* expect to see differences between "january" and "february", or are we more interested in the "upwelling season" versus the "non upwelling season"?
glm_11 <- glmer(neighbor ~ location + upwell + (1|filename), data=glm_data, family="binomial")
summary(glm_11) #upwelling season did not affect interactions between cory and urchins, but location was significant 

binned_residuals(glm_11) #75% residuals within error bounds
testDispersion(glm_11) #good dispersion
```



```{r}
#visualizing
library(car)
library(MASS)
library(ggplot2)
library(psych)

glm_data2 <- subset(glm_data, select = c("year", "neighbor")) 

write.csv(glm_data2, "glm_data2.csv")
glm_data2 <- read.csv("glm_data2.csv")
#View(glm_data2)

#graph that works, stand_enso does not work as y 
ggplot(glm_data2, aes(x=year, y=neighbor)) + geom_smooth(method = "glm", method.args = list(family = "binomial")) 
```

Amelia's addition - plot % interactions over time by location
```{r}
#create df with total number of rows for each location at each year
glm_n <- glm_data %>%
  group_by(year, location) %>% 
  summarize(n=n()) %>% 
  ungroup()

#create df with number of positive (1) interactions for each location at each year
glm_1 <- glm_data %>%
  group_by(year, location) %>%
  filter(neighbor=="1") %>% 
  summarize(sum=sum(neighbor)) %>% 
  ungroup()
  
#combine the two df and calculate percentages
glm_percent <-  merge(glm_1, glm_n, by=c("year", "location")) %>% 
  mutate(percent = sum/n) #calculate percent interactions

#standard dev and error to make error bars 
SD = sd(glm_percent$percent)
SE = sd(glm_percent$percent)/sqrt(length(glm_percent$percent))

glm_percent$sd <- sd(glm_percent$percent)
glm_percent$se <- sd(glm_percent$percent)/sqrt(length(glm_percent$percent))
#View(glm_percent)

pd <- position_dodge(0.9) 

#woohoo! now the data are ready to be plotted
ggplot(glm_percent, aes(x=year, y=percent, fill=location)) +
  geom_bar(position="dodge", stat="identity")+ geom_errorbar(aes(x=year, ymin=percent-SE, ymax=percent+SE),width=0.4, position=pd) #error bars all the same size? 

#also, here's a nifty function to save ggplots!
ggsave('figures/amelia-plot.png', width = 4, height = 2.5)
```


```{r}
#making results presentable 
#install.packages("huxtable")
#install.packages("jtools")
library(huxtable)
library(jtools)
summ(glm4)
summ(glm5)

library(sjPlot)
tab_model(glm4)#this table is a good way to summarize the significant predeictors of the model 
tab_model(glm5)
```


```{r}
#next: error bars- how to get different ones for each bar 
```
